Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-11-26 Paper/application.tex
finish fix reviews

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that previously supported only basic string operations: concatenation and regular constraints. The original CertiStr was incapable of verifying any XSS attack due to the lack of support for transducers and replacement operations.
CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.

@@

The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a \textbf{one-hour} time limit per test. Table~\ref{tab:string_operations} summarizes the results. The results from CertiStrR were classified into three categories: SAT (satisfiable), UNSAT (unsatisfiable), and Inconclusive. An "Inconclusive" result indicates that the solver cannot determine satisfiability due to the inherent incompleteness of the string solver's algorithm, not timeout (all tests completed within one hour). This occurs when the solver cannot determine whether string variables need to be assigned different values that satisfy all constraints.

@@

>>>>

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that previously supported only basic string operations: concatenation and regular constraints. The original CertiStr was incapable of verifying any XSS attack due to the lack of support for transducers and replacement operations.
CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.
The OCaml code of the SFT implementation is extracted from our Isabelle formalization using Isabelle's extraction mechanism, ensuring that the core solving algorithm of CertiStr remains certified.

@@

The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a \textbf{one-hour} time limit per test. Table~\ref{tab:string_operations} summarizes the results. The results from CertiStrR were classified into three categories: SAT (satisfiable), indicating that there exist assignments of string variables satisfying the constraints; UNSAT (unsatisfiable), indicating that no such assignments exist; and Inconclusive. An "Inconclusive" result indicates that the solver cannot determine satisfiability due to the inherent incompleteness of the string solver's algorithm, not timeout (all tests completed within one hour). This occurs when the solver cannot determine whether string variables need to be assigned different values that satisfy all constraints.

@@

Our approach demonstrates superior efficiency compared to other string solving approaches for several key reasons. First, our SFTs are implemented with compact symbolic labels that enable efficient representation of infinite alphabets, and we employ red-black trees for storing states, ensuring logarithmic-time lookup and insertion operations. Second, our forward propagation strategy proves more efficient than alternative approaches such as term rewriting or backpropagation employed by other solvers. While forward propagation supports only a limited set of string constraints—for instance, it is only complete for string constraints satisfy tree properties \cite{cpp/KanLRS22}—this trade-off between completeness and efficiency enables our solver to handle large-scale benchmarks that remain intractable for more general but computationally expensive approaches.

### 2025-11-26 Paper/formalization.tex
finish fix reviews

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'i,'b)"$| NFT |$\Rightarrow$|
   |$('q,'a)$| NFA |$\Rightarrow$||$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}$|
   |$\Rightarrow 'b\text{ set option})\Rightarrow$| |$('q\times 'q, ~'b)$| eNFA where
"productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
  |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
  |$\Delta_e =$| (* case of |$\epsilon$|-transitions *)
      |$\{((p,p'),~\texttt{the }(((\mathcal{M}_t~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid $|
       |$p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A}~\land$| |$(p, (\texttt{None}, f), q)\in \Delta_t~\mathcal{T}$|
        |$ \land~\exists S.~(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{Some } S\} ~\cup$|
      (* case of non-|$\epsilon$|-transitions *)
      |$\{((p,p'),~\texttt{the }(\texttt{F}~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),(q, q'))\mid $|
       |$p, p', q, \sigma_1, \sigma_2, q', f.~(p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land$|
       |$ (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
          |$ \exists S.~\texttt{F}~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
  |$\Delta_e' =$| (* case of |$\epsilon$|-transitions *)
      |$\{((p,p'),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
       |$(p, (None, f), q)\in \Delta_t~\mathcal{T}~\land $|
       |$(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
      (* case of non-|$\epsilon$|-transitions *)
      |$\{((p,p'),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
        |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}$|
        |$\land~\sigma_1\cap \sigma_2 \neq \emptyset~\land\exists x\in(\sigma_1\cap\sigma_2).$|
        |$~((\mathcal{M}_t~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
  |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
  |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of product operation}
\label{fig-def-FTProd}
\end{figure}

@@

>>>>

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('{q},~'a,~'i,~'b)"$| NFT
  |$\Rightarrow$| |$('q,~'a)$| NFA |$\Rightarrow$||$(('a,~'b)$| Tlabel |$\Rightarrow~'a \text{ set}$|
  |$\Rightarrow~'b\text{ set option})\Rightarrow$| |$('q~\times~'q, ~'b)$| eNFA where
"productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
  |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
  |$\Delta_e =$| (* case of |$\epsilon$|-transitions *)
      |$\{((p,p'),~\texttt{the }(((\mathcal{M}_t~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid $|
       |$p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A}~\land$| |$(p, (\texttt{None}, f), q)\in \Delta_t~\mathcal{T}$|
        |$ \land~\exists S.~(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{Some } S\} ~\cup$|
      (* case of non-|$\epsilon$|-transitions *)
      |$\{((p,p'),~\texttt{the }(\texttt{F}~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),(q, q'))\mid $|
       |$p, p', q, \sigma_1, \sigma_2, q', f.~(p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land$|
       |$ (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
          |$ \exists S.~\texttt{F}~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
  |$\Delta_e' =$| (* case of |$\epsilon$|-transitions *)
      |$\{((p,p'),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
       |$(p, (None, f), q)\in \Delta_t~\mathcal{T}~\land $|
       |$(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
      (* case of non-|$\epsilon$|-transitions *)
      |$\{((p,p'),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
        |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}$|
        |$\land~\sigma_1\cap \sigma_2 \neq \emptyset~\land\exists x\in(\sigma_1\cap\sigma_2).$|
        |$~((\mathcal{M}_t~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
  |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
  |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of product operation}
\label{fig-def-FTProd}
\end{figure}

@@

The formalization of SFA/SFT presents distinct challenges compared to classical FA/FT, primarily stemming from the symbolic nature of transition labels. Unlike classical automata where labels are concrete alphabet symbols, SFAs and SFTs use Boolean algebras to represent potentially infinite label sets. This introduces two key complications. First, our formalization must accommodate different Boolean algebra implementations (intervals, bit vectors, predicates, etc.) while maintaining uniform semantics, necessitating the locale-based abstraction layer that separates the fixed automata structure from variable label theories. Second, operations on transition labels—such as intersection and membership testing—require Boolean algebra computations rather than simple syntactic comparisons. This demands explicit reasoning about semantic preservation: we must prove that operations on Boolean algebra representations (e.g., interval intersection) correctly preserve the set semantics defined by the \texttt{sem} function. These complications, absent in classical automata where label operations are trivial, required careful architectural design and substantial additional proof effort to establish correctness.
